Fast Sorting on Modern Computers

### 2005-08-14 docs/tech_report/elementary.tex
fixed up the references
changes for testing
fixed all references

\code{Selection sort code}{selectsort.c}
\subsection{Testing}
%TESTING
All elementary sorts are only sorted up to 65536 keys in the simulations, and
262144 keys with the performance counters. Due to the quadratic nature of
these sorts, days and weeks would be required to run tests on the larger arrays.

@@

\label{selection branches logn}
The branch prediction performance is not expected to be bad. Flow
control predictions are very straightforward, and should result in
very few misses.  Comparison predictions, however, are very
numerous. Each traversal of the array has an average of $N/2$
comparisons, as we search for the smallest key in the unsorted part of
the array. During the search, the algorithm compares each key with the
smallest key seen so far. In a random array, we would expect that the
candidate for the smallest key (the left-to-right minimum) would
change frequently towards the start of the array, making the
comparison branch rather unpredictable. However, as the traversal
continues, the left-to-right minimum will become smaller, and thus
more difficult to displace. It will become very unusual to find a
smaller key, and thus the branch will almost always resolve in the
same direction.  Therefore, the comparison branch will quickly become
very predictable.

Knuth \cite{Knuth1} analyses the number of changes to the
left-to-right minimum when searching for the minimum element of an
array. He shows that for $N$ elements, the number of changes is $H_N -
1$, where $H_N$ is the harmonic series:

@@

%TODO check that this is put in after the titlepage
%TODO check that the captions are the same as the captions at the end of the
%other chapters
% I'm doing a slightly different result here. Level 2 doesn't interest us, so put
% in 2 prediction results instead - 1 for select/insert, 1 for bubble/shaker
\afterpage{
\label{Elementary sort results}

>>>>

\code{Selection sort code}{selectsort.c}
\subsection{Testing}

All elementary sorts are only sorted up to 65536 keys in the simulations, and
262144 keys with the performance counters. Due to the quadratic nature of
these sorts, days and weeks would be required to run tests on the larger arrays.

@@

\label{selection branches harmonics}
The branch prediction performance is not expected to be bad. Flow
control predictions are very straightforward, and should result in
very few misses.  Comparison predictions, however, are very
numerous. Each traversal of the array has an average of $N/2$
comparisons, as we search for the smallest key in the unsorted part of
the array. During the search, the algorithm compares each key with the
smallest key seen so far. In a random array, we would expect that the
candidate for the smallest key (the left-to-right minimum) would
change frequently towards the start of the array, making the
comparison branch rather unpredictable. However, as the traversal
continues, the left-to-right minimum will become smaller, and thus
more difficult to displace. It will become very unusual to find a
smaller key, and thus the branch will almost always resolve in the
same direction.  Therefore, the comparison branch will quickly become
very predictable.

Knuth \cite{Knuth97} analyses the number of changes to the
left-to-right minimum when searching for the minimum element of an
array. He shows that for $N$ elements, the number of changes is $H_N -
1$, where $H_N$ is the harmonic series:

@@

% I'm doing a slightly different result here. Level 2 doesn't interest us, so put
% in 2 prediction results instead - 1 for select/insert, 1 for bubble/shaker
\afterpage{
\label{Elementary sort results}

### 2005-08-14 docs/tech_report/heap.tex
fixed all references

\section{A more detailed look at branch prediction}

@@

More importantly, the accuracy of the each successive comparison increases in
the same way as selection sort (see Section \ref{selection branches logn}),
becoming more predictable as the fanout increases. For example, in the 8-heap,
shown in Figures \ref{base and memory-tuned heapsorts-3} and \ref{base and
memory-tuned heapsorts-4}, the first comparison is predicted correctly 50\% of
the time, while the next is predicted correctly 61\% of the time, then 71\%,
77\%, and so on in a curve, which is shown in Figure \ref{memory-tuned heapsort
curve}.

>>>>

\section{A More Detailed Look at Branch Prediction}

@@

More importantly, the accuracy of the each successive comparison increases in
the same way as selection sort (see Section \ref{selection branches harmonics}),
becoming more predictable as the fanout increases. For example, in the 8-heap,
shown in Figures \ref{base and memory-tuned heapsorts-3} and \ref{base and
memory-tuned heapsorts-4}, the first comparison is predicted correctly 50\% of
the time, while the next is predicted correctly 61\% of the time, then 71\%,
77\%, and so on in a curve, which is shown in Figure \ref{memory-tuned heapsort
curve}.

### 2005-08-14 docs/tech_report/quick.tex
fixed all references

\section{Branch predictor results}

\subsection{A more detailed look at branch prediction}

@@

Figure \ref{base, cache and multi quicksorts}(a) shows the behaviour
of the indiviual branches in base quicksort. The $i$ branch is the
loop control branch for the index sweeping from left-to-right up the
array during partition. The $j$ branch is the corresponding loop
sweeping down the array. The partition step terminates when $i \geq
j$, and the $partition end$ branch checks that condition. The
$insertion$ branch is the comparison branch in the insertion sort
algorithm. Recall that our quicksort algorithm does not sort
partitions which are smaller than 10 elements, but instead does a
final pass of insertion sort over the whole array to sort these small
partitions. Finally, the $median$ branch is the sum of the results for
the branches involved in computing the median-of-three for each
partition.

@@

When we look at multi-quicksort, there is a 19\% drop in the number of
iterations of the $i$ and $j$ loops, with a corresponding reduction in
the number of mispredictions. Interestingly, the percentage
misprediction rate of the $i$ and $j$ branches does not change --- it
is 37\% in both cases.

@@

Multi-quicksort does, however, introduce a lot of additional
branches. We already saw in Figure \ref{quick instructions}, that the
binary search and sequential search variants increase the number of
executed instructions by around 25\% and 50\% respectively. Figure
\ref{base, cache and multi quicksorts} shows additional measures of the $binary
left$ and $binary right$ branches, which are the branches within the
binary search code. The number of binary search branches is much
greater than the reduction in $i$ and $j$ branches. Furthermore, these
binary search branches are highly (43\%) unpredictable.

@@

\label{quicksort entropy}
When we originally looked at quicksort, we expected that the $i$ and
$j$ branches in quicksort would resolve in each direction. In fact,
this is not the case. The main reason is that the chosen pivot is
almost never the true median of the section of the array to be
partitioned. Thus, the $i$ and $j$ branches are almost always biased,
which makes them predictable. According to Knuth, if the pivot element
is chosen randomly, then the $i$ and $j$ branches will be taken (i.e.
the $i$ or $j$ loop will execute) twice as often as they are not taken
(i.e. the loop exits).

In order to investigate this, we implemented a simple version of
quicksort that uses the middle element as the pivot rather than a
median-of-three\footnote{It is important to note that selecting
the pivot based on only a single element can be particularly inefficient
because it removes the natural sentinels at the top and bottom
of the regions to be partitioned that we get from median-of-three. In
Figure \ref{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9} we see that this results in an additional branch to handle
the edgecase without a sentinel. Although this branch is almost 100\%
predictable, it is executed a very large number of times.}. The arrays we sort are random, so the value of this
element will also be random. Figure \ref{base quicksort without a median, with median of 3, and with pseudo medians of 5, 7 and 9}(a)
shows the behaviour of each of the important branches in this code. We
see that the $i$ and $j$ branches are indeed taken about twice as
often as not-taken, and the $i$ and $j$ branches are correctly
predicted around 71\% of the time.

The result that the branches are more predictable (71\%) than they are
biased (66\%) is a particularly interesting finding. The reason is that
the bias in the results is an average figure over a large number of
partitions that quicksort performs to sort the array. On a particular
partition, the bias will be more pronounced because the $i$ and $j$
branches will be biased in opposite directions.

For example, consider the case where the pivot element is larger than
the true median. In this case, the final resting place of the pivot
will be towards the right-hand end of the region to be
partitioned. Thus, the $i$ loop will execute a large number of times,
and thus the $i$ loop will be biased towards being taken. The $j$
loop, on the other hand, will be biased toward not-taken, because the
average element is likely to be smaller than the pivot. Therefore, the
$j$ loop will execute only a small number of times. Thus, we see that
in a particular partition, we do not normally get an equal number of
executions of the $i$ and $j$ loops. Instead we get a large number of
executions in one loop (which makes its branch relatively
predictable), whereas the other loop often does not enter at all
(which makes its branch also predictable, although executed a much
smaller number of times). Thus, although overall the $i$ and $j$
branches are taken twice as often as not-taken, on a particular
partition one will usually be biased in the not-taken direction, and
as a result will be executed a relativelty small number of times.

In the preceding paragraphs we have seen that the $i$ and $j$ branches
are predictable when we have a pivot that is larger or smaller than
the true median of the array section to be partitioned. An interesting
question is the effect of choosing a better approximation of the true
median on branch prediction. Figure \ref{base quicksort without a
median, with median of 3, and with pseudo medians of 5, 7 and 9}(b)
shows the behavior of these branches when median-of-three is used to
select the pivot. Perhaps the biggest effect is that the number of
executions of the $i$ and $j$ branches falls by around 15\%. This fall
is partly offset by the comparison branches to compute the median, but
these account for only around 1.25\% of the original branches from the
no-median quicksort. Thus, there is a reduction of almost 14\% in the
number of executed comparison branches. Even more interesting, however,
is the bias in the remaining $i$ and $j$ branches. The no-median $i$ and
$j$ branches are 66\% biased, but the median-of-three branches are
only 60\% biased in the taken direction. The accuracy with which
these branches are correctly predicted falls from 71\% to 64\%.

The reason for the reduction in bias and prediction accuracy is that
median-of-three gives a much better estimate of the true median of the
section to be partitioned. The final resting place of the pivot will
usually be closer to the middle of the section, to there is less bias
in the branches. Thus, although median-of-three gives around a 14\%
reduction in executed comparison branches, each of those branches is
much less likely to be correctly predicted. Overall there is actually
a 6\% \textit{increase} in the total number of branch mispredictions.

>>>>

\section{A More Detailed Look at Branch Prediction}

@@

Figure \ref{base, memory-tuned and multi- quicksorts-1} shows the behaviour of
the indiviual branches in base quicksort. The \cc{i} branch is the loop control
branch for the index sweeping from left-to-right up the array during partition.
The \cc{j} branch is the corresponding loop sweeping down the array. The
partition step terminates when $i \geq j$, and the \cc{partition end} branch
checks that condition. The \cc{insertion} branch is the comparison branch in the
insertion sort algorithm. Recall that our quicksort algorithm does not sort
partitions which are smaller than 10 elements, but instead does a final pass of
insertion sort over the whole array to sort these small partitions. Finally, the
\cc{median} branch is the sum of the results for the branches involved in
computing the median-of-three for each partition.

@@

When we look at multi-quicksort, there is a 19\% drop in the number of
iterations of the \cc{i} and \cc{j} loops, with a corresponding reduction in the
number of mispredictions. Interestingly, the percentage misprediction rate of
the \cc{i} and \cc{j} branches does not change --- it is 37\% in both cases.

@@

Multi-quicksort does, however, introduce a lot of additional branches. We
already saw in Figure \ref{quick instructions}, that the binary search and
sequential search variants increase the number of executed instructions by
around 25\% and 50\% respectively. Figure \ref{base, memory-tuned and multi-
quicksorts-3} shows additional measures of the \cc{binary left} and \cc{binary
right} branches, which are the branches within the binary search code. The
number of binary search branches is much greater than the reduction in \cc{i}
and \cc{j} branches. Furthermore, these binary search branches are highly (43\%)
unpredictable.

@@

\label{quicksort entropy}
When we originally looked at quicksort, we expected that the \cc{i} and \cc{j}
branches in quicksort would resolve in each direction. In fact, this is not the
case. The main reason is that the chosen pivot is almost never the true median
of the section of the array to be partitioned. Thus, the \cc{i} and \cc{j}
branches are almost always biased, which makes them predictable. According to
Knuth, if the pivot element is chosen randomly, then the \cc{i} and \cc{j}
branches will be taken (i.e.  the \cc{i} or \cc{j} loop will execute) twice as
often as they are not taken (i.e. the loop exits).

In order to investigate this, we implemented a simple version of quicksort that
uses the middle element as the pivot rather than a median-of-three\footnote{It
is important to note that selecting the pivot based on only a single element can
be particularly inefficient because it removes the natural sentinels at the top
and bottom of the regions to be partitioned that we get from median-of-three. In
Figure \ref{quick-predictors2} we see that this results in an additional branch
to handle the edgecase without a sentinel. Although this branch is almost 100\%
predictable, it is executed a very large number of times.}. The arrays we sort
are random, so the value of this element will also be random. Figure \ref{base
quicksort without a median, with median of 3, and with pseudo-medians of 5, 7
and 9-1} shows the behaviour of each of the important branches in this code. We
see that the \cc{i} and \cc{j} branches are indeed taken about twice as often as
not-taken, and the \cc{i} and \cc{j} branches are correctly predicted around
71\% of the time.

The result that the branches are more predictable (71\%) than they are biased
(66\%) is a particularly interesting finding. The reason is that the bias in the
results is an average figure over a large number of partitions that quicksort
performs to sort the array. On a particular partition, the bias will be more
pronounced because the \cc{i} and \cc{j} branches will be biased in opposite
directions.

For example, consider the case where the pivot element is larger than the true
median. In this case, the final resting place of the pivot will be towards the
right-hand end of the region to be partitioned. Thus, the \cc{i} loop will
execute a large number of times, and thus the \cc{i} loop will be biased towards
being taken. The \cc{j} loop, on the other hand, will be biased toward
not-taken, because the average element is likely to be smaller than the pivot.
Therefore, the \cc{j} loop will execute only a small number of times. Thus, we
see that in a particular partition, we do not normally get an equal number of
executions of the \cc{i} and \cc{j} loops. Instead we get a large number of
executions in one loop (which makes its branch relatively predictable), whereas
the other loop often does not enter at all (which makes its branch also
predictable, although executed a much smaller number of times). Thus, although
overall the \cc{i} and \cc{j} branches are taken twice as often as not-taken, on
a particular partition one will usually be biased in the not-taken direction,
and as a result will be executed a relativelty small number of times.

In the preceding paragraphs we have seen that the \cc{i} and \cc{j} branches are
predictable when we have a pivot that is larger or smaller than the true median
of the array section to be partitioned. An interesting question is the effect of
choosing a better approximation of the true median on branch prediction. Figure
\ref{base quicksort without a median, with median of 3, and with pseudo-medians
of 5, 7 and 9-2} shows the behavior of these branches when median-of-three is
used to select the pivot. Perhaps the biggest effect is that the number of
executions of the \cc{i} and \cc{j} branches falls by around 15\%. This fall is
partly offset by the comparison branches to compute the median, but these
account for only around 1.25\% of the original branches from the no-median
quicksort. Thus, there is a reduction of almost 14\% in the number of executed
comparison branches. Even more interesting, however, is the bias in the
remaining \cc{i} and \cc{j} branches. The no-median \cc{i} and \cc{j} branches
are 66\% biased, but the median-of-three branches are only 60\% biased in the
taken direction. The accuracy with which these branches are correctly predicted
falls from 71\% to 64\%.

The reason for the reduction in bias and prediction accuracy is that
median-of-three gives a much better estimate of the true median of the section
to be partitioned. The final resting place of the pivot will usually be closer
to the middle of the section, to there is less bias in the branches. Thus,
although median-of-three gives around a 14\% reduction in executed comparison
branches, each of those branches is much less likely to be correctly predicted.
Overall there is actually a 6\% \n{increase} in the total number of branch
mispredictions.

### 2005-08-14 docs/tech_report/shell.tex
fixed all references

%TODO
\section{Predictor Results}

@@

%TODO see figure something
In shellsort, almost exactly half of the important branches are mispredicted.
Simply predicting that all branches are taken would result in a slightly higher
prediction ratio of 50.8\%. Similarly, for improved shellsort, slightly more
than half of the branches are correctly predicted, at 50.3\%, rising to 53\%
using a strategy of predicting all branches as taken.

@@

The expected number of branch mispredictions is $log_{(\frac{11}{5})}N$. In
actual fact, it is almost exactly correct, being slightly lower (18.48
mispredictions per key, instead of 19.34, for 4194304 keys).

>>>>

%TODO
\section{A More Detailed Look at Branch Prediction}

@@

%TODO see figure something
In shellsort, almost exactly half of the important branches are mispredicted.
Simply predicting that all branches are taken would result in a slightly higher
prediction ratio of 50.8\%. Similarly, for improved shellsort, slightly more
than half of the branches are correctly predicted, at 50.3\%, rising to 53\%
using a strategy of predicting all branches as taken. These results can be seen
in Figures \ref{Branch prediction performance for Shellsorts}.

@@

The expected number of branch mispredictions is $log_{(\frac{11}{5})}N$.  This
is almost exactly correct, being slightly lower (18.48 mispredictions per key,
instead of 19.34, for 4194304 keys).
